---
title: "Introduction to MaMaMIA"
author: "Aleksey Ermolaev"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to MaMaMIA}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  markdown: 
    wrap: 72
---

```{r knit-param, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>",
    fig.width = 8,
    fig.height = 2.5,
    dpi = 96,
    out.width = "100%"
)
```

# Overview

**MaMaMIA** (Max Mapping for Meticulous Introgression Analysis) detects
introgressed segments from Genotyping-by-Sequencing (GBS) coverage data.
The workflow has five main steps:

1.  Prepare windowed donor and recipient coverage and GC-content tables.
2.  Build an RCA (Read Count Array) from those tables.
3.  Correct coverage for GC-content bias.
4.  Segment pairwise donor-recipient coverage differences with circular
    binary segmentation (CBS).
5.  Inspect putative introgressions and optionally export them as BEDPE.

This vignette walks through that pipeline using the packaged `triticum`
example dataset (*Triticum timopheevii* donor versus *Triticum aestivum*
Chinese Spring T2T recipient).

```{r setup}
library(MaMaMIA)
```

# Preparing input data

MaMaMIA expects windowed coverage (mapped read counts) and GC-content
tables for both the **donor** and **recipient** genomes. The same
preprocessing is applied independently to each genome. Downstream
comparison matches windows by position index, not by shared genomic
coordinates, so every chromosome-donor and recipient-must be tiled into
the **same number of windows**.

## Competitive mapping

Reads are mapped to a concatenated reference that merges the donor and
recipient genomes into one FASTA. Competitive mapping assigns each read
to the genome (and chromosome) where it aligns best, so coverage
reflects genome-specific signal rather than ambiguous cross-mapping
between related sequences. More background is given in [Feuerborn et al.
(2020)](https://doi.org/10.1186/s12864-020-07229-y).

Mapped reads are filtered with `samtools` to retain primary, mapped
alignments (`-F 2308`) of sufficient mapping quality (e.g.
$\mathrm{MAPQ} \ge 20$).

## Windowed coverage and GC content

Filtered BAMs are summarised in fixed genomic windows (we recommend
100-1000 windows per chromosome, but it is up to you) with `bedtools`.
For a chromosome length file in [Genome
format](https://bedtools.readthedocs.io/en/latest/content/general-usage.html#genome-file-format)
and a filtered BAM `sample.bam`:

``` bash
bedtools makewindows -g chrom.genome -n 1000 > windows.bed
bedtools nuc -fi concatenated_genomes.fa -bed windows.bed | cut -f1-3,5 > sample.gc
bedtools coverage -a windows.bed -b sample.bam -counts > sample.cov
```

`bedtools nuc` reports several columns; column 5 is the fraction of G/C
bases in the window. Coverage and GC tables passed to `RCA()` must each
have exactly four columns (names are ignored):

| Column | Meaning                                       |
|--------|-----------------------------------------------|
| 1      | Chromosome ID                                 |
| 2      | Start (0-based)                               |
| 3      | End (exclusive)                               |
| 4      | Value (read count or GC fraction in $[0, 1]$) |

Within each genome, coverage and GC tables must use identical window
coordinates. On each chromosome, windows must form contiguous half-open
intervals in row order (no gaps or overlaps). A chromosome metadata
table (`meta`) supplies three columns: chromosome ID, a short display
name, and a subgenome label. Donor and recipient chromosome IDs must be
disjoint, as must their subgenome labels.

# Load example data

```{r load-data}
data(triticum)
```

`triticum` provides precomputed tables from this pipeline:

- `don_cov`, `don_gc` — *T. timopheevii* coverage and GC (14
  chromosomes; At and G subgenomes; 1000 windows each)
- `rec_cov`, `rec_gc` — *T. aestivum* Chinese Spring T2T coverage and GC
  (21 chromosomes; A, B, and D subgenomes; 1000 windows each)
- `meta` — chromosome ID, short name (e.g. `"1At"`, `"1A"`), and
  subgenome for all 35 chromosomes

# Build an RCA object

`RCA()` joins donor and recipient windowed coverage and GC-content with
chromosome metadata into a Read Count Array. It validates tiled windows,
matching coverage/GC coordinates within each genome, equal window counts
across all chromosomes, and disjoint donor/recipient chromosome IDs and
subgenome labels. The result starts with `corrected = FALSE`. Set
`presorted = FALSE` only if input tables are not already ordered by
chromosome and coordinates (bedtools output usually is).

```{r create-rca}
RCA.object <- RCA(
    dcov = don_cov,
    dgc = don_gc,
    rcov = rec_cov,
    rgc = rec_gc,
    meta = meta
)
summary(RCA.object)
```

`print()` reports chromosome counts (total, donor, recipient), window
count, and GC-correction status. `summary()` lists chromosomes per
subgenome and whether correction has been applied (`corrected = FALSE`
at this stage).

# Correct GC bias

`correctReadCounts()` marks outlier windows, fits a zero-inflated
negative binomial (ZINB) model of coverage as a smooth function of GC
content (with subgenome as a covariate), and stores GC-corrected
coverage in `cor.gc`. Ideal windows (retained after outlier filtering)
are used for model fitting and later for segmentation. Outlier cutoffs
are fixed at the 1%/99% GC quantiles and the 99% coverage quantile.
Increase `cores` to speed up `glmmTMB` fitting on larger datasets:

```{r correct-gc}
corrected_data <- correctReadCounts(RCA.object, cores = 4L)
summary(corrected_data)
```

After correction, `summary()` also reports the GC and coverage bounds
stored in `$outliers`, and marks `corrected = TRUE`. The fitted model is
kept in `$fit`.

## Resource use

Fitting differentiates the likelihood window by window, so both runtime and peak
memory scale with the number of windows that are fitted, and not with `cores`.
Budget roughly **0.1 MB per window** (about 1 GB per 10,000 windows on top of a
small baseline) and around 0.85 s per 1,000 windows at `cores = 4`. This
walkthrough fits 34,000 windows and peaks near 3.6 GB, so fitting is practical up
to a few tens of thousands of windows on a typical machine, while 350,000 windows
would need roughly 36 GB. Raising `cores` shortens the fit substantially without
changing the memory it needs.

Fitted values also shift by around `1e-7` relative with `cores`, so a batch is
better run as **one process per sample with `cores = 1`** than as one wide
multi-threaded call. Results are then independent of the thread count, so a given
sample gives the same answer whichever worker runs it; peak memory stays bounded
per worker, and throughput comes from the number of workers rather than from
threads, for example with `mirai` or `crew`, or a shell loop over samples with
`OMP_NUM_THREADS=1` exported in each worker's environment.

```{r batch-fits, eval = FALSE}
# In each worker, with OMP_NUM_THREADS=1 exported in its environment:
corrected <- correctReadCounts(rca, cores = 1L)
```

## Plotting coverage profiles

`plot()` on an RCA object can show raw (`"orig_cov"`) or GC-corrected
(`"corr_cov"`) coverage. By default only ideal windows are drawn; set
`show_outliers = TRUE` to include filtered windows. If correction has
not been run, `plot()` warns and falls back to raw coverage with
outliers shown.

```{r original-cov-plot}
plot(corrected_data, plot.type = "orig_cov")
```

```{r corrected-cov-plot}
plot(corrected_data, plot.type = "corr_cov")
```

```{r corrected-cov-plot-outliers}
plot(corrected_data, plot.type = "orig_cov", show_outliers = TRUE)
```

## Zero-inflated negative binomial (ZINB) model

Let $Y_i$ be the coverage in window $i$ with GC content $\mathrm{GC}_i$
and subgenome $S_i$. Then

$$Y_i \sim \mathrm{ZINB}(\mu_i, \theta, \pi_i)$$

where

$$\mu_i = f(\mathrm{GC}_i) + \beta_{S_i}, \qquad
\operatorname{logit}(\pi_i) = g(\mathrm{GC}_i) + \gamma_{S_i}$$

with $f$ and $g$ fitted as smooth splines. Corrected coverage scales the
observed count by the ratio of expected coverage at a reference GC
(median GC of ideal windows) to the fitted expectation at the window's
own GC.

# Subset coverage data to selected subgenomes or chromosomes

Before segmentation, `subset()` can restrict an RCA object to selected
donor and/or recipient chromosomes. Specify targets by subgenome label,
by chromosome ID, or both (the two filters are combined as a union on
each side). At least one argument must be non-`NULL`; omitting filters
for one side leaves that side unchanged. The result must retain at least
one donor and one recipient chromosome.

By subgenome:

```{r subset-subgenome-plot}
subset_cd <- subset(
    corrected_data,
    don_subgenomes = "At",
    rec_subgenomes = "A"
)
plot(subset_cd, plot.type = "corr_cov")
```

By chromosome ID:

```{r subset-chromosome-plot, eval = FALSE}
subset(
    corrected_data,
    don_chromlist = c("OY997261.1", "OY997263.1"),
    rec_chromlist = c("CP155614.1", "CP155611.1")
)
```

The rest of this vignette uses the subgenome-level subset (`At` / `A`)
above.

# Reversing chromosome orientation

When the donor and recipient assemblies are built independently,
homologous chromosomes may be assembled in opposite orientations. Their
per-window coverage and GC vectors then run in opposite directions along
the genome, so position-by-position comparison aligns the wrong windows.

`reverseWindows()` flips the effective orientation of selected
chromosomes so the two genomes can be compared directly. The window
indexes (`iid`, `start`, `stop`, and the chromosome identifier) are
kept unchanged; instead the value vectors (coverage, GC-corrected
coverage, and the per-window validity flags) are reversed within each
selected chromosome. Chromosomes are selected by their **ID** (the
`chr_id` column of the meta table), not by their short display name
(`chr_name`):

```{r reverse-rca}
# Reverse donor chromosome 1At
subset_cd_rev1At <- reverseWindows(
    subset_cd,
    chr_ids = "OY997261.1"
)
# Compare with original 1At chromosome
plot(subset_cd_rev1At, plot.type = "corr_cov")
```

This orientation marker is preserved when the chromosome is passed on:
segmentation (`segments()`, below) and plotting see the reversed
profile.

# Segmentation of donor-recipient difference coverage

For each donor-recipient chromosome pair, `segments()` computes the
difference in GC-corrected coverage on aligned ideal windows and runs
circular binary segmentation (CBS) via
[`DNAcopy::segment()`](https://rdrr.io/bioc/DNAcopy/man/segment.html).
CBS recursively partitions the difference profile into contiguous
regions of constant mean ([Olshen et al.
2004](https://doi.org/10.1093/biostatistics/kxh008); [Venkatraman &
Olshen 2007](https://doi.org/10.1093/bioinformatics/btl646)). The result
is an `ISA` (Introgression Segment Array) object.

## Donor-recipient chromosome pairs

Segmentation requires an explicit named vector `target_pairs`: names are
donor chromosome IDs and values are recipient chromosome IDs. There is
no default all-pairs expansion; only the listed pairs are analysed. Here
each donor chromosome is paired with the corresponding recipient
chromosome from the subset:

```{r segments}
don_chromosomes <- subset_cd$meta$don_chr_ids
rec_chromosomes <- subset_cd$meta$rec_chr_ids
target_pairs <- setNames(rec_chromosomes, don_chromosomes)
aat_segments <- segments(subset_cd,
    target_pairs = target_pairs
)
aat_segments
```

A single pair can be specified directly:

```{r segments-one-pair, eval = FALSE}
segments(
    subset_cd,
    target_pairs = c("OY997261.1" = "CP155614.1")
)
```

## CBS parameters

CBS behaviour is controlled by three arguments passed through to
`DNAcopy::segment()` (with `undo.splits = "sdundo"`). Defaults are
`alpha = 0.001`, `min_width = 2`, and `undo.SD = 1`.

- `alpha` — significance level for the change-point test. Smaller values
  accept fewer splits and yield coarser segments (`0 < alpha <= 1`).
- `min_width` — minimum number of windows retained after segmentation;
  shorter segments are dropped. Note that `DNAcopy::segment()` itself is
  always called with `min.width = 2`; this value is applied as a
  post-filter (`min_width >= 2`).
- `undo.SD` — number of standard deviations between adjacent segment
  means required to keep a split. Larger values prune more splits
  attributed to local trends (`0 < undo.SD <= 10`).

Custom parameters:

```{r segments-custom-param, eval = FALSE}
segments(
    subset_cd,
    target_pairs = target_pairs,
    alpha = 0.01,
    min_width = 5,
    undo_SD = 1
)
```

## Introgression detection thresholds

After CBS, each segment is summarised by its mean (`seg.mean`) and
median (`seg.median`) coverage difference. A segment is flagged as a
putative introgression when the chosen statistic is at or above a
threshold:

- `threshold_type` — `"median"` (default) or `"mean"`.
- `min_segment_median` — threshold when `threshold_type = "median"`
  (default `0`).
- `min_segment_mean` — threshold when `threshold_type = "mean"` (default
  `0`).

With the defaults, any segment whose median difference is non-negative
is marked as a putative introgression. Raise the threshold to require a
stronger donor excess, or switch to the mean:

```{r segments-threshold, eval = FALSE}
# Stricter median threshold
segments(
    subset_cd,
    target_pairs = target_pairs,
    min_segment_median = 5,
    threshold_type = "median"
)

# Use segment mean instead of median
segments(
    subset_cd,
    target_pairs = target_pairs,
    threshold_type = "mean",
    min_segment_mean = 5
)
```

These flags can be revised later with `subset()` on the `ISA` object
without re-running CBS (see `?subset.ISA`).

## Reversing after segmentation

`reverseWindows()` also operates on an `ISA` object. When a chromosome
included in the pairwise comparison is reversed, the difference profile
of every donor-recipient pair containing it changes, so segmentation is
**recomputed** for those pairs automatically. Segments of pairs that do
not involve a reversed chromosome are left untouched.

```{r reverse-isa, eval = FALSE}
aat_segments <- reverseWindows(
    aat_segments,
    chr_ids = "OY997261.1"
)
```

Apply the reversal as early as possible - on the `RCA` object before
`segments()` - and only use the `ISA` form when you need to flip a
chromosome in an already-segmented result.

# Plotting of segmentation result

`plot()` on an `ISA` object draws the pairwise profiles already stored
in that object (the pairs chosen at `segments()` time). Two styles are
available via `plot.type`.

## Mirrored coverage (`plot.type = "mirror"`)

Donor GC-corrected coverage is plotted above zero and recipient coverage
below, so shared structure and local excesses are easy to compare along
the chromosome:

```{r plot-mirror}
plot(aat_segments, plot.type = "mirror")
```

## Coverage difference (`plot.type = "diff"`)

The per-window difference is shown with a zero reference line.
Horizontal bars mark the segment summary used for introgression calling
(`seg.median` or `seg.mean`, as recorded in
`aat_segments$meta$metrics`). By default only segments flagged as
putative introgressions are drawn; set `all_segments = TRUE` to overlay
every CBS segment:

```{r plot-diff}
plot(aat_segments, plot.type = "diff")
```

```{r plot-diff-all, eval = FALSE}
plot(aat_segments, plot.type = "diff", all_segments = TRUE)
```

To restrict the plot to a subset of pairs, filter the `ISA` object first
with `subset()` (see `?subset.ISA`), then call `plot()` on the result.

# Inspecting putative introgressions

`summary()` on an `ISA` object lists each putative introgression with
donor and recipient chromosome, subgenome, start, end, and length.
Coordinates can be reported in `bp`, `Kb`, `Mb`, or `Gb`:

```{r summary-isa}
summary(aat_segments, unit = "Mb", digits = 2)
```

Thresholds and pairs can be adjusted without re-running CBS. For
example, keep only one pair and require a higher median difference:

```{r subset-isa, eval = FALSE}
aat_strict <- subset(
    aat_segments,
    target_pairs = c("OY997261.1" = "CP155614.1"),
    min_segment_median = 5
)
summary(aat_strict, unit = "Mb")
```

# Export putative introgressions to BEDPE

`writeToBedpe()` writes segments with `putative_introgression = TRUE` to
a tab-separated BEDPE file. Each row is one donor–recipient interval
pair. Columns include standard BEDPE fields plus chromosome names and
subgenome labels; `score` is the active segment metric (`seg.median` or
`seg.mean`):

```{r export-bedpe}
bedpe_file <- tempfile(fileext = ".bedpe")
writeToBedpe(aat_segments, file = bedpe_file)
read.table(bedpe_file, header = TRUE, sep = "\t")
```

If no putative introgressions remain after filtering, a header-only file
is written and a warning is issued.

# Session info

```{r session-info}
sessionInfo()
```
